Generating Interactive WebSocket Applications in TypeScript

### 2020-03-08 PLACES_2020/casestudy.tex
PLACES: address reviewers/FZ comments

\begin{figure}
\begin{lstlisting}[language=JavaScript, tabsize=4]
const handleP1Move: S13 = (move: Point) => {
	board.P1(move);			// User logic
	if (board.won()) { return [Labels.Lose, move, move]; }
	else if (board.draw()) { return [Labels.Draw, move, move]; }
	else { return [Labels.Update, move, handleP2Move]; }}

@@

\paragraph{Game Server}
We set up the WebSocket server as an Express.js \cite{ExpressJS}
application on top of a Node.js \cite{NodeJS} runtime.
We define our own game logic in a \texttt{Board} class to keep track of the
game state and expose methods to query the result.
This custom logic is integrated into our \texttt{handleP1Move} and
\texttt{handleP2Move} handlers (\cref{lst:svrprotocol}), so the session runtime
can handle \texttt{Pos(Point)} messages from players and transition according
to the injected game logic.

Note that, by TypeScript's structural typing \cite{TypeScriptSpec}, replacing
\texttt{handleP2Move} on line 5 with \texttt{handleP1Move} would be
type-correct -- this allows for better code reuse as opposed to defining
additional abstractions to work around the limitations of nominal typing in
\cite{Hybrid2016}.
There is also full type erasure when transpiling to JavaScript to run the
server code, so state space explosion is not a runtime consideration.

@@

For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses}
uses \textit{higher-order components} (HOC) to build the correct state
implementations depending on which player the user chooses to be.
We leverage the \textit{Redux} \cite{Redux} state management library to keep track of game
state, thus showing the flexibility of our library design in being
interoperable with other libraries and idiomatic JavaScript practices.

>>>>

\begin{figure}
\begin{lstlisting}[language=JavaScript, tabsize=4]
const handleP1Move: S13 = (move: Point) => {
	board.P1(move);			// User logic
	if (board.won()) { return [Labels.Lose, move, move]; }
	else if (board.draw()) { return [Labels.Draw, move, move]; }
	else { return [Labels.Update, move, move, handleP2Move]; }
}

@@

\paragraph{Game Server}
We set up the WebSocket server as an Express.js \cite{ExpressJS}
application on top of a Node.js \cite{NodeJS} runtime.
We define our own game logic in a \texttt{Board} class to keep track of the
game state and expose methods to query the result.
This custom logic is integrated into our \texttt{handleP1Move} and
\texttt{handleP2Move} handlers (\cref{lst:svrprotocol}), so the session runtime
can handle \texttt{Pos(Point)} messages from players and transition to the 
permitted successor states (\cref{lst:game}) according
to the injected game logic: if \texttt{P1} played a winning move (line 3), 
\texttt{Svr} sends a \texttt{Lose} message to \texttt{P2} with the winning 
move, and also sends a \texttt{Win} message to \texttt{P1} 
(note that we do not need to specify the label for a single choice); 
if \texttt{P1}'s move resulted 
in a draw (line 4), \texttt{Svr} sends \texttt{Draw} messages to 
both \texttt{P2} and \texttt{P1}; otherwise, the game continues (line 5), 
so \texttt{Svr} updates both \texttt{P2} and \texttt{P1} with the latest 
move and proceeds to handle \texttt{P2}'s turn.

Note that, by TypeScript's structural typing \cite{TypeScriptSpec}, replacing
\texttt{handleP2Move} on line 5 with a recursion of \texttt{handleP1Move} would be
type-correct -- this allows for better code reuse as opposed to defining
additional abstractions to work around the limitations of nominal typing in
\cite{Hybrid2016}.
There is also full type erasure when transpiling to JavaScript to run the
server code, so state space explosion is not a runtime consideration.

@@

For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses}
uses \textit{higher-order components} (HOC) to build the correct state
implementations depending on which player the user chooses to be.
We leverage the \textit{Redux} \cite{Redux} state management library to keep track of game
state, thus showing the flexibility of our library design in being
interoperable with other libraries and idiomatic JavaScript practices.
% REVIEW: tight-coupling?
Our approach encourages the separation of concerns between the 
communication logic and program logic -- the generated session runtime
keeps track of the state of the EFSM to ensure protocol conformance by 
construction, whilst \textit{Redux} solely manages our game state.

### 2020-03-08 PLACES_2020/encoding.tex
PLACES: make wrapfig float to avoid overflow
PLACES: address reviewers/FZ comments

Our session runtime performs communication in a protocol-conformant manner, but
does not expose these IO actions to the developer by delegating the
aforementioned responsibilities to an inner class.
The runtime listens to message (receive) events on the communication channel,
invokes the corresponding callback to obtain the value to send next, and
performs the send.
The developer instantiates the session by constructing an instance of the
session runtime class, providing the WebSocket endpoint URL (to open the
connection) and initial state (to execute the EFSM).

@@

\begin{wrapfigure}{r}{0.5\textwidth}
  \begin{center}
    \includegraphics[width=0.5\textwidth]{figures/efsm_p1.png}
  \end{center}

@@

For browser-side targets, we extend the approach presented in \cite{MVU2019} on
\textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU)
architecture.
Each state in the EFSM is a \emph{model type} and uniquely defines a
\textit{view function}, set of \textit{messages} and \textit{update function}:
the view function defines what to render to the DOM, the set of messages define
the possible (IO) actions available at that state, and the update function
defines which successor state to transition to, given some supported IO action
at this state.

>>>>

The generated code for our session runtime performs communication in a protocol-conformant manner, but
does not expose these IO actions to the developer by delegating the
aforementioned responsibilities to an inner class.
% REVIEW: it is not clear what is
% generated on the server-side and what actually is the session runtime
The runtime executes the EFSM by keeping track of 
the current state (similar to the generated code in \cite{javatypestate}) and only permitting the allowed IO actions at the current state.
The runtime listens to message (receive) events on the communication channel,
invokes the corresponding callback to obtain the value to send next, and
performs the send.
The developer instantiates the session by constructing an instance of the
session runtime class, providing the WebSocket endpoint URL (to open the
connection) and initial state (to execute the EFSM).

@@

\begin{wrapfigure}{R}{0.5\textwidth}
  \begin{center}
    \includegraphics[width=0.5\textwidth]{figures/efsm_p1.png}
  \end{center}

@@

For browser-side targets, we extend the approach presented in \cite{MVU2019} on
\textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU)
architecture.
% FEEDBACK: introduce MVU and model type
An MVU application features a \textit{model} encapsulating application 
state, a \textit{view function} rendering the state on the DOM, and an 
\textit{update function} handling \textit{messages} produced by the 
rendered model to produce a new model.
The concept of model types express type dependencies between these 
components: a \emph{model type} uniquely defines a \textit{view function}, 
set of \textit{messages} and \textit{update function} -- rather than 
producing a new model, the update function defines valid transitions to 
other model types.
We leverage the correspondence between model types and states in the EFSM:
each state in the EFSM is a model type, the set of messages represent
the possible (IO) actions available at that state,
and the update function defines which successor state to transition to, 
given the supported IO actions at this state.

### 2020-03-08 PLACES_2020/intro.tex
PLACES: address reviewers/FZ comments

\fz{This paragraph needs to be modified, to emphasis that our generated APIs
are correct by construction, and not ``verifying'' existing ones}
We see the application of the MPST methodology to verifying interactive web
applications written in TypeScript to be an interesting design space -- to what
extent can the MPST methodology be applied such that we leverage TypeScript
language features to maximise static safety guarantees whilst preserving a
flexible, natural and idiomatic workflow for TypeScript developers?
Such a workflow would ultimately decrease the overhead for incorporating MPST
into mainstream web development, which reduces development time by programmatically
verifying communication correctness of the implementation.

>>>>

We see the application of the MPST methodology to generating interactive 
TypeScript web applications to be an interesting design space 
-- to what extent can the MPST methodology be applied to
deliver a workflow where developers use the generated TypeScript APIs in
their application to guarantee protocol conformance by construction?
Such a workflow would ultimately decrease the overhead for incorporating MPST
into mainstream web development, which reduces development time by programmatically
verifying communication correctness of the implementation.
